---
title: "Exercise: Monte Carlo Integration"
subtitle: "An interactive review"
author: "Alexander Massey"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    number-sections: true
    code-fold: false
    code-copy: true
    code-overflow: wrap
    highlight-style: github
    theme: flatly
    fontsize: 1.05em
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
---


```{r}
#| label: setup
#| include: false

library(ggplot2)
library(dplyr)

set.seed(2026)
```

# Introduction

Monte Carlo integration is a way to approximate a definite integral using
random sampling.  It turns an integration problem into an estimation problem: draw random
points, evaluate the function at those points, and average. 

As the sample size grows, the **Law of Large Numbers (LLN)** tells us that the estimate converges to the true value and the **Central Limit Theorem (CLT)** tells us the asymptotic distribution of our estimate based on a finite sample size.

In this exercise, you'll be provided a smooth wiggly function to estimate.  You will plot it, make an estimate of its integral by taking a random sample, and build a confidence interval for your estimate.  At the end of the exercise, we will check by simulation whether that confidence interval behaves the way the CLT promises it should.

# The Monte Carlo Integral

Suppose we want to estimate

$$
I = \int_a^b f(x)\, dx.
$$

If $x$'s are drawn independently from a $\text{Uniform}(a,b)$
distribution, then the $f(x)$'s are i.i.d. random variables with

$$
E[f(x)] = \int_a^b f(x)\,\left(\frac{1}{b-a}\right)\, dx = \frac{I}{b-a}.
$$

Rearranging, the natural **Monte Carlo estimate** of $I$ is

$$
\hat{I}_n = (b-a) \cdot \frac{1}{n}\sum_{x \in s} f(x).
$$

In words: the integral is (width of the interval) $\times$ (average height of
the function). By the LLN, $\hat{I}_n \to I$
as $n \to \infty$.

# The Function

We'll work with the function

$$
f(x)=1+2\cdot\sin(\frac{x}{2})\cdot e^{-x/6}
$$
In a new R script, copy---paste this function in and put it into memory (highlight it and press ctr/cmd + enter)
```{r}
#| label: define-f

f <- function(x) 1 + 2 * sin(x / 2) * exp(-x / 6)
```

Our objective is to integrate $f(x)$ over the interval $x\in[1, 9]$.

## Visualize $f(x)$

Before integrating anything, get a feel for the function. Plot $f(x)$ over
$[1, 9]$.

::: {.callout-tip collapse="true"}
## Hint 1 — Plotting the curve

``` r
x <- seq(from = 0, to = 10, by = 0.1)
y <- f(x)
plot(x, y, type = "l", lwd = 4, col = "steelblue",
     xlab = "x", ylab = "f(x)", 
     xlim = c(0,10), ylim = c(0, max(y)),
     main = "f(x) = 1 + 2 sin(x/2) exp(-x/6)")
abline(v = c(1,9), lwd = 5, lty = 2)
```
:::

**Questions to consider:**

1. Is $f(x)$ positive over the whole interval? What would happen if $f(x)$ was positive and negative in our intregation domain?
2. Roughly by eye, what do you think the average height of the curve is over
   $[1,9]$?  \textbf{Pro-tip:} Even if it is just a guess, always keep intuition about \text{what makes sense} before, during and after you compute anything.

# Analytical Integration

Before turning to Monte Carlo, it's worth knowing the *true* value of the
integral so we have something to check our simulations against.

## The analytical solution

Unlike many integrals in real life, this one is simple enough to do by hand, but the point here is not to do it yourself.  Hence the antiderivative is provided below.

$$
F(x) = x - \frac{6}{5}e^{-x/6}\sin\!\left(\frac{x}{2}\right) - \frac{18}{5}e^{-x/6}\cos\!\left(\frac{x}{2}\right) + C,
$$

so that

$$
I = \int_1^9 f(x)\,dx = F(9) - F(1).
$$

The R code below implements $F(x)$ directly and confirms it matches it with R's native numerical integration function `integrate()`:

```{r}
#| label: analytical-solution

F <- function(x) {
  x - (6/5) * exp(-x/6) * sin(x/2) - (18/5) * exp(-x/6) * cos(x/2)
}

I_analytical <- F(9) - F(1)
I_analytical

I_numeric <- integrate(f, lower = 1, upper = 9)$value
I_numeric

all.equal(I_analytical, I_numeric)
```

The true value is approximately **`r round(I_analytical, 4)`**. Keep this
number handy — it's the benchmark for everything that follows.

# Monte Carlo Estimation

Now let's approximate the same integral with sampling.

## Sample and estimate

1. Draw `n = 1000` random values from a $\text{Uniform}(1, 9)$ distribution.
2. Evaluate $f$ at each sampled value.
3. Combine these into the Monte Carlo estimate $\hat{I}_n = (b-a)\cdot \overline{f}$.
4. Compare your estimate to `I_analytical`. How close is it?

::: {.callout-tip collapse="true"}
## Hint 3 — Sampling and estimating

``` r
n <- 1000
a <- 1
b <- 9

x_sample <- runif(n, min = a, max = b)
f_sample <- f(x_sample)

I_hat <- (b - a) * mean(f_sample)
I_hat
```

Try re-running this chunk a few times. Notice that `I_hat` changes slightly each time — it's a random
quantity, unlike the analytical value.
:::

**Question:** If you increase `n` from 1000 to 100,000, what do you expect to
happen to `I_hat`?

# Quantifying Uncertainty: The CLT and LLN

The LLN tells us $\hat{I}_n$ converges to $I$, but it says nothing about how
far off we should expect to be for a *finite* $n$. That's where the CLT comes
in.

Because different draws of $f(x)$ are i.i.d. with some finite variance
$\sigma^2 = \mathbb{V}(f(x))$, the CLT tells us that for large $n$,

$$
\overline{f} \; \dot\sim \; N\!\left(\frac{I}{b-a},\; \frac{\sigma^2}{n}\right).
$$

Multiplying through by $(b-a)$, the Monte Carlo estimate itself is
approximately normal:

$$
\hat{I}_n \;\dot\sim\; N\!\left(I,\; (b-a)^2\frac{\sigma^2}{n}\right).
$$

We estimate $\sigma^2$ with the sample variance of our draws of $f(x)$,
which gives an approximate **95% confidence interval**:

$$
\hat{I}_n \pm 1.96 \cdot (b-a) \cdot \frac{s}{\sqrt{n}}, \qquad
s = \sqrt{\mathbb{V}[f(x)]}.
$$
In R we can estimate $s$ with the functions `sd()` or by taking the square root (i.e. `sqrt`) of the sample variance `\var()`.  

\mathbf{Pro-tip:} The z-score can be retrieved dynamically in R by using the function `qnorm(1-(alpha/2))` after defining `alpha` as the crtical level $\alpha$. 

## Draw a sample and build a 95% CI

Using a sample of size `n = 1000`:

1. Compute the point estimate $\hat{I}_n$.
2. Compute the standard error of the estimate.
3. Construct a 95% confidence interval.
4. Does your interval contain the analytical value from Exercise 2?

::: {.callout-tip collapse="true"}
## Hint 4 — Standard error and CI

``` r
n <- 1000
x_sample <- runif(n, 1, 9)
f_sample <- f(x_sample)

I_hat <- (9 - 1) * mean(f_sample)
se_hat <- (9 - 1) * sd(f_sample) / sqrt(n)

alpha <- 0.05
z <- qnorm(1-(alpha/2))

ci_lower <- I_hat - z * se_hat
ci_upper <- I_hat + z * se_hat

c(estimate = I_hat, lower = ci_lower, upper = ci_upper)
```
:::

**Question:** One single confidence interval either contains the true value
or it doesn't — the true value is fixed and the CI is random. So what does "95% confidence" actually describe? Keep this question in mind for the next section, where you'll check it directly.

# Validating the CLT: Empirical Coverage

If the CLT-based confidence interval is doing its job, then if we repeated
the sampling process many times and built a new 95% CI each time, about 95%
of those intervals should contain the true value $I$ — regardless of the
sample size (though small $n$ can make the \textit{asymptotically} normal approximation shakier).

## Coverage simulation

For each of three sample sizes — say `n = 10`, `n = 50`, and `n = 500` —
we will repeat the following many times (e.g., `n_reps = 300`):

1. Draw a sample of size $n$ from $\text{Uniform}(1,9)$.
2. Compute $\hat{I}_n$ and its 95% CI.
3. Record whether the interval contains the true value `I_analytical`.

Then we will:

- Compute the **empirical coverage rate** for each sample size (the
  proportion of intervals that contained the truth). It should be close to
  0.95 for all three, though small-$n$ coverage may wobble a bit more.
- Make a plot with:
  - a horizontal line at the true value $I$,
  - every simulated CI drawn as a thin, light-gray vertical segment,
  - the segments jittered/spread along the x-axis so they don't all
    overlap, and
  - the three sample sizes shown separately (facets or distinct x-ranges
    both work).

Copy---paste the following simulation code.

::: {.callout-tip collapse="false"}
## Simulation function

``` r
one_ci <- function(n, a = 1, b = 9) {
  x_sample <- runif(n, a, b)
  f_sample <- f(x_sample)
  I_hat <- (b - a) * mean(f_sample)
  se_hat <- (b - a) * sd(f_sample) / sqrt(n)
  data.frame(
    lower = I_hat - 1.96 * se_hat,
    upper = I_hat + 1.96 * se_hat
  )
}

n_reps <- 300
sample_sizes <- c(10, 50, 500)

sims <- lapply(sample_sizes, function(n) {
  reps <- do.call(rbind, replicate(n_reps, one_ci(n), simplify = FALSE))
  reps$n <- n
  reps$rep_id <- 1:n_reps
  reps$covered <- (reps$lower <= I_analytical) & (I_analytical <= reps$upper)
  reps
})

sims_df <- do.call(rbind, sims)
head(sims_df)
```
:::

::: {.callout-tip collapse="false"}
## Calculate coverage rate (percentage of time CI contains the true value)

``` r
coverage <- sims_df |>
  group_by(n) |>
  summarise(coverage_rate = mean(covered), .groups = "drop")

coverage
```

Each `coverage_rate` should land close to 0.95. For lower $n$ the asymptotic validity of the CLT may not have "kicked in".  Try switching out `qnorm()` with the score from the t-distribution`qt(1-(alpha/2), df = n-1)`
:::

::: {.callout-tip collapse="false"}
## Plotting the intervals

``` r
ggplot(sims_df, aes(x = rep_id, y = (lower + upper) / 2)) +
  geom_segment(
    aes(xend = rep_id, y = lower, yend = upper,
        color = covered),
    linewidth = 0.4,
    alpha = 0.6
  ) +
  scale_color_manual(
    values = c(`TRUE` = "gray70", `FALSE` = "firebrick"),
    labels = c(`TRUE` = "Covers truth", `FALSE` = "Misses truth"),
    name = NULL
  ) +
  geom_hline(
    yintercept = I_analytical,
    color = "steelblue",
    linewidth = 1
  ) +
  facet_wrap(
    ~ n,
    scales = "free_x",
    labeller = label_both
  ) +
  labs(
    x = "Simulation replicate (jittered)",
    y = "Estimate and 95% CI",
    title = "Empirical coverage of Monte Carlo confidence intervals",
    subtitle = "Steelblue line = true integral value"
  ) +
  theme_minimal()
```

You could also color the segments by whether they covered the true value
(`covered`) to make the small fraction of misses visually obvious.
:::

**Reflection questions:**

1. Look at your plot. The horizontal blue line — the true value $I$ — never
   moves. What *is* random from replicate to replicate?
2. Does the coverage rate get noticeably better as $n$ increases, or is it
   already close to 95% even at `n = 10`? What does this tell you about how
   quickly the CLT approximation "kicks in" for this particular $f$?
3. Suppose someone tells you, "there's a 95% probability that the true
   integral lies inside my one confidence interval." Using what you just
   saw in the simulation, explain why that phrasing is misleading, and how
   you'd rephrase it correctly.
4. How would you expect the *width* of the CIs (not their coverage) to
   change as $n$ grows from 10 to 500? Does your plot support that?

# Wrap-Up

- The Monte Carlo estimate $\hat{I}_n = (b-a)\overline{f}$ is consistent
  by the LLN — it converges to the true integral as $n \to \infty$.
- The CLT lets us attach a confidence interval to any *finite* sample,
  because $\hat{I}_n$ is approximately normally distributed around $I$.
- Confidence intervals themselves are random estimates, not the probability that the true value is in that interval (it is or it isn't because it is fixed).


------------------------------------------------------------------------
