---
title: "Model-assisted estimation"
subtitle: "Estimating Mean Basal Area in the Doubs"
author: "Alexander Massey"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    number-sections: true
    code-fold: false
    code-copy: true
    code-overflow: wrap
    highlight-style: github
    theme: flatly
    fontsize: 1.05em
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

```{r}
#| label: setup
#| include: true

packages <- c("sf", "terra", "tidyterra", "dplyr", "ggplot2", "tmap", "here", "tidyverse", "leaflet", "shiny")

installed <- rownames(installed.packages())
to_install <- setdiff(packages, installed)

if (length(to_install) > 0) {
  install.packages(to_install)
}

library(sf)
library(terra)
library(tidyterra)
library(dplyr)
library(ggplot2)
library(tmap)
library(tidyverse)
library(leaflet)
library(shiny)
library(here)

```

# Introduction

In this exercise you'll work with basal area (BA) data from the Doubs
department in France. ONF (Office National des Forêts) measured basal area
on a set of field plots, but **only in public forests**, which make up
roughly a third of the forest land in Doubs. Separately, IGN produced a
wall-to-wall kNN prediction of basal area by combining photogrammetric
canopy height information and the official French National 
Forest Inventory (NFI), covering the whole department.  Being a survey with
an independent sample, this provides a perfect opportunity to demonstrate the 
difference between true external models and internal models in the context of 
model-assisted estimation which is design-based.

You will build up several estimators of the population mean basal area,
starting from basic one-phase estimation (no auxiliary information) and 
gradually bringing in the auxiliary
prediction surface, and compare their point estimates, variances, and
confidence intervals. Along the way you'll think carefully about **which of
these estimators are actually unbiased**, and under what assumptions.

# The Estimators

Let $Y(x)$ be the basal area measured on field plot $x\in s$, for sample $s$,
and let $Z(x)$ be the value of the auxiliary surface at that
same plot location. Let $\bar Z := \frac{1}{A_T}\int_T Z(x)dx$ denote the mean of the 
auxiliary over the *entire* population (every pixel) in target territory $T$.  This is
called the **exhaustive mean**.  

## One-phase estimator

If we ignore the prediction surface entirely, we're back to a simple
design-based estimator:

$$
\hat{Y}_{1p} = \frac{1}{n}\sum_{x\in s} Y(x), \qquad
\hat{V}(\hat{Y}_{1p}) = \frac{1}{n(n-1)}\sum_{x\in s}\left(Y(x)-\bar{Y}_{s}\right)
$$

These are exactly unbiased under the assumed sampling design, but it uses none of the
auxiliary information we have available.  For systematic sampling, simulation studies 
have shown that the point estimator can still be considered unbiased, but the variance
estimator tends to overestimate the true design-based variance.  However, we generally proceed
as if it is uniform i.i.d. sampling and accept that the confidence interval is overly conservative.

## Synthetic estimator

At the other extreme, for wall-to-wall auxiliary information this is the mean of some model's
predictions applied to $Z(x)$ for every pixel on a map.  Since there are so many pixels on a map, this can
be considered be negligly close the notion under the Monte Carlo approach:

$$
\hat{Y}_{syn} = \frac{1}{A_T}\int \widehat{Y}(x)dx
$$
where $A_T$ is the area of the target region.  There is no bias correction, no guarantee
that it is unbiased at all, but can be applied to any region where we have remote sensing,
even if it has zero inventory plots containing ground truth.  Sometimes this is the best you 
can do.

**Question to consider:** If you computed the "variance" of this estimator
across the millions of pixels in the prediction surface, what would happen
to it, and why would that number be misleading?

## Difference estimator

The **difference estimator** corrects the synthetic estimate using the
average residual observed on the field plots:

$$
\hat{Y}_{diff} = \frac{1}{A_T}\int \widehat{Y}(x)dx + \frac{1}{n}\sum_{x\in s} (R(x)),
\quad
\hat{V}(\hat{Y}_{diff}) = \frac{1}{n(n-1)}\sum_{x\in s}\left(R(x)-\bar{R}_{s}\right)
$$
where $R(x):=Y(x)-\hat{Y}(x)$.

This estimator treats the prediction surface as fixed, external
information — it was **not** fit using the field plots in our target area and rather some
independent training set.

**Question to consider:** Assuming uniform, independent sampling, is
$\hat{V}(\hat{Y}_{diff})$ exactly unbiased, asymptotically unbiased, or
biased? Does that still hold if the plots were instead laid out on a
systematic grid?

## Model-assisted (plug-in) estimator

Instead of taking the prediction surface as-is, we can **re-fit** it using
the field data — for example, a simple linear model:

$$
Y(x) = \beta_0 + \beta_1 Z(x) + R(x)
$$
In this exercise we consider $Z(x)$ to be the synthetic kNN prediction from an indepedent
survey (the official NFI) so it can be **considered as if was another predictor**, like say mean canopy height.  

Because ordinary least squares residuals have mean zero by construction for the training set, we
can plug the exhaustive mean of the predictor directly into the fitted line:

$$
\hat{Y}_{plugin} = \frac{1}{A_T}\int\left(\hat\beta_0 + \hat\beta_1\,{Z(x)}\right)dx, \quad
\widehat{V}(\hat{Y}_{plugin}) \approx \frac{1}{n(n-1)}\sum_{x\in s}\left(\hat R(x)-\bar{\hat{R}}_{s}\right).
$$
where $\hat R(x)$ are the estimated residuals of the linear model fit.  The hat
indicates that it was estimated from the probability sample.
A useful heuristic for how much this should help, relative to the one-phase
estimator, comes from the model's $R^2$:

$$
\hat{V}(\hat{Y}_{plugin}) \approx (1 - R^2)\, \hat{V}(\hat{Y}_{1p}).
$$

**Question to consider:** Why did we not need to separately compute a mean
residual correction here, the way we did for the difference estimator? Write
what the plug-in estimator would look like if we chose a model without the 
zero mean residual property.

::: {.callout-tip collapse="true"}
## Hint
$$
\begin{array}{c @{\hspace{2cm}} l}
\begin{aligned}
Y(x):= \mathbf{Z}(x)^t\boldsymbol{\beta} + R(x)
\end{aligned}
& \text{\textcolor{blue}{\textit{"assisting model"} proposal}} \\[10pt]
\begin{aligned}
\hat{Y}(x) = \mathbf{Z}(x)^t\boldsymbol{\hat\beta}_s
\hspace{1.5cm}
\hat{R}(x) = Y(x) - \hat{Y}(x)
\end{aligned}
& \text{\textcolor{blue}{\textit{Plug-in to }}} \textcolor{blue}{\hat{Y}_{diff}}
\end{array}
$$
$$
\widehat{Y}_{plugin}=\frac{1}{A_T}\int \hat Y(x)dx + \frac{1}{n}\sum_{x\in s}\hat R(x)
$$
:::

## g-weight variance

The variance formula above for $\hat{Y}_{plugin}$ may be considered a bit unrealistic
— it ignores the fact that $\hat\beta_0$ and $\hat\beta_1$ were themselves
estimated from the same sample, with their own sampling variability. The
**g-weight variance** accounts for this properly using a sandwich-type
covariance estimate of the regression coefficients, propagated through the
exhaustive auxiliary vector.

$$
\hat{Y}_{gw}=\bar{\mathbf Z}^\top\mathbf{\hat{\boldsymbol{\beta}}_s},
\qquad\qquad
\hat{\mathbb V}[\hat{Y}_{gw}] = 
  \displaystyle
  \bar{\mathbf Z}^\top
  \hat\Sigma_{\hat{\boldsymbol{\beta}}_s}
  \bar{\mathbf Z}
$$
where $\bar{\mathbf Z}$ is the exhaustive auxiliary vector over all pixels and the design-based
variance-covariance estimate of $\hat{\boldsymbol{\beta}}_s$ is
$$
\hat\Sigma_{\hat{\boldsymbol{\beta}}_s}
\approx
\mathbf{A}_s^{-1}
\left[
\frac{1}{n^2}
\sum_{x \in s}
\hat{R}(x)^2
\mathbf{Z}(x)\mathbf{Z}(x)^t
\right]
\mathbf{A}_s^{-1}
$$
with $\mathbf{A}_s:=\frac{1}{n}\sum_{x \in s}\mathbf{Z}(x)\mathbf{Z}(x)^\top$.

This variance estimator has better statistical properties (i.e. calibration) and its
derivation is based on a Taylor approximation, which is out of scope for this course.

Regardless, you'll build this one from scratch later in the exercise.

# The Data

You'll work with four spatial layers:

- `doubs_ONF_plots.gpkg` — field plots measured by ONF, with basal area
  stored in `BA_ONF`, and a logical column `uniform_sample` flagging a
  systematic subsample of plots.
- The ONF public forest boundary shapefile — the extent ONF actually
  inventories (roughly a third of the forest area in Doubs).
- `doubs_strata.gpkg` — a stratification layer, with region codes stored in
  `REGIONN`.
- `kNN_predictive_surface_doubs.tif` — the wall-to-wall kNN prediction of basal
  area from NFI data, stored in band `BA_pred_NFI`.

```{r}
#| label: load-data
#| echo: true
#| output: false
#| message: false

doubs_ONF_plots <- st_read(here("exercises", "data", "doubs_ONF_plots.gpkg")) |>
  mutate(BA_ONF = G_onf,
         surf_maille_fct = as.factor(surf_maille))

doubs_public_forests <- st_read(here("exercises", "data", "doubs_public_forests.gpkg"))

doubs_strata <- st_read(here("exercises", "data", "doubs_strata.gpkg"))

pred_surface <- rast(here("exercises", "data", "kNN_predictive_surface_doubs.tif"))
```

## Visualize the layers

Before doing anything quantitative, look at what you're working with. Use
the **zoom** R package (or repeated `ggplot2` calls) to get a feel for the
extent and shape of each layer, then overlay the field plots on the
stratification layer, colored by `REGIONN`.

::: {.callout-tip collapse="true"}
## Hint 1 — Plots over strata with ggplot()

``` r
ggplot() +
  geom_sf(data = doubs_strata, aes(fill = REGIONN), color = NA, alpha = 0.6) +
  geom_sf(data = doubs_ONF_plots, size = 0.7, shape = 16) +
  theme_minimal() +
  labs(title = "ONF Plots over Stratification Regions", fill = "REGIONN")
```
:::

::: {.callout-tip collapse="true"}
## Interactive plot to zoom in using the *tmap* package

``` r
# Interactive map
tmap_mode("view")

# comment in/out uniform sample part
tm_shape(doubs_strata) +
  tm_polygons(
    col = "REGIONN",
    fill_alpha = 0.5,
    border.col = "grey40"
  ) +
  tm_shape(doubs_ONF_plots) +
  tm_symbols(
    col = "surf_maille_fct",
    size = 0.4,
    fill_alpha = 0.9,
    border.col = "black"
  )
```
:::

**Questions to consider:**

1. Some `REGIONN` strata clearly have more plots per unit area than others —
   ONF has *oversampled* them in part for LiDAR calibration. What might other 
   reasons to oversample? Think about what makes a stratum expensive or important 
   to monitor more precisely.
2. `surf_maille` gives the mesh size of the sampling grid, and
   `uniform_sample` flags a subsample that restores a uniform mesh size
   across all of Doubs. Re-plot using only `doubs_ONF_plots |> filter(uniform_sample)`, 
   again over `REGIONN`. How does the pattern change?  
3. Is the uniform sample still purely systematic?

::: {.callout-tip collapse="true"}
## Hint 2 — Uniform subsample only

``` r
ggplot() +
  geom_sf(data = doubs_strata, aes(fill = REGIONN), color = NA, alpha = 0.6) +
  geom_sf(data = doubs_ONF_plots |> filter(uniform_sample),
          size = 0.7, shape = 16) +
  theme_minimal() +
  labs(title = "Uniform-Mesh Subsample over Stratification Regions",
       fill = "REGIONN")
```
:::

## Restricting the population to public forest

ONF only inventories **public** forest — about a third of the forest land
in Doubs. Any population mean we compute needs to be defined over that same
extent, so intersect the stratification layer with the public forest
boundary.

```{r}
#| label: intersect-frame

doubs_strata <- st_transform(doubs_strata, st_crs(doubs_public_forests))

doubs_intersected <- st_intersection(
  doubs_public_forests,
  doubs_strata |> select(REGIONN)
)

# Clean up: keep one polygon per row
doubs_intersected <- doubs_intersected |>
  st_collection_extract("POLYGON") |>
  st_cast("POLYGON")

doubs_intersected$area_ha <- as.numeric(st_area(doubs_intersected)) / 10000
area_total <- sum(doubs_intersected$area_ha)
```

**Question to consider:** Basal area is normally reported as a per-hectare
mean, not a total.  Why would we want to compute `area_ha` at all?  Give an 
example of situation where the total of a parameter of interest would be desirable 
to report.

## Building the exhaustive auxiliary vector

To apply the difference and model-assisted estimators, we need
$\bar{\mathbf{Z}} =\frac{1}{A_T}\int \mathbf{Z}(x)dx$: the mean of the prediction surface **restricted to public forest**, since that's the population ONF actually samples.  I have already done this for you to reduce the file size of `kNN_predictive_surface_doubs.tif` in the git repo but this is the code I used if you are interested...

::: {.callout-tip collapse="true"}
# This is how to crop the raster if you need to...
```{r}
#| eval: false
#| label: exhaustive-mean

doubs_vect <- vect(doubs_intersected)

if (crs(doubs_vect) != crs(pred_surface)) {
  doubs_vect <- project(doubs_vect, crs(pred_surface))
}

crop(pred_surface, doubs_vect, mask = TRUE)
```
:::

Calculate the exhaustive vector
```{r}
# Calculate the exhaustive vector
exhaustive_vector <- global(pred_surface, fun = "mean", na.rm = TRUE)
```

Now join the plots to the public forest polygons so you can plot only the
plots that actually fall inside the extent you're estimating over.

```{r}
#| label: join-plots

doubs_plots_on_layers <- st_join(
  doubs_ONF_plots,
  doubs_public_forests,
  join = st_within,
  left = TRUE
)
```

## Map the prediction surface with sample plots

Overlay the uniform-mesh subsample of plots on the cropped prediction
surface.

::: {.callout-tip collapse="true"}
## Hint 3 — Plotting the raster with plots

``` r
ggplot() +
  geom_spatraster(data = pred_surface$BA_pred_NFI) +
  geom_sf(data = doubs_plots_on_layers |> filter(uniform_sample),
          color = "red", size = 1, shape = 16) +
  scale_fill_viridis_c(na.value = "transparent", name = "Prediction") +
  theme_minimal() +
  labs(title = "kNN Predictive Surface with ONF Plots")
```
:::

**Question to consider:** Many maps express forests as large continuous areas.  Is this forest layer like this or more fractal-like. 

# Estimation

For all of the estimators below, use the uniform-mesh subsample so that
each plot represents roughly the same area.

```{r}
#| label: prep-sample

sample_plots <- doubs_plots_on_layers |> filter(uniform_sample)
n <- nrow(sample_plots)

alpha <- 0.05
z_score <- qnorm(c(alpha/2, 1-alpha/2))
```

## One-phase estimate

Compute $\hat{Y}_{1p}$, its variance, and a 95% confidence interval based on
the CLT.

::: {.callout-tip collapse="true"}
## Hint 4 — One-phase

``` r
Y_1p  <- mean(sample_plots$BA_ONF)
VY_1p <- var(sample_plots$BA_ONF) / n
CI_Y_1p <- Y_1p + z_score * sqrt(VY_1p)
```
:::

## Synthetic estimate

Compute the synthetic estimate two ways: first as the mean of predictions
*at the plot locations*, then compare it conceptually to the exhaustive
mean `exhaustive_vector`.

**NOTE:** In this data, we were not provided the training data from the official
NFI used to produce the kNN predictions so we can't properly propagate 
the randomness to produce a confidence interval.  Typically for kNN this is done
using a bootstrap applied to all the wall-to-wall pixels. Be aware that the CIs 
for the synthetic estimator are often narrower than the model-assisted ones, which
leads to getting *greedy* and desiring to trust it more.

::: {.callout-tip collapse="true"}
## Hint 5 — Synthetic estimator

``` r
Y_syn  <- as.numeric(exhaustive_vector)
VY_syn <- NA
CI_Y_syn <- Y_syn + z_score * sqrt(VY_syn)
```
:::

**Question to consider:** Is the synthetic estimator exactly unbiased,
approximately unbiased, or biased? What would have to be true about the
model for it to be unbiased?

## Difference estimate

Compute $\hat{Y}_{diff}$ and its confidence interval.

::: {.callout-tip collapse="true"}
## Hint 6 — Difference estimator

``` r
Y_diff  <- as.numeric(exhaustive_vector) +
  mean(sample_plots$BA_ONF - sample_plots$BA_pred_NFI)
VY_diff <- var(sample_plots$BA_ONF - sample_plots$BA_pred_NFI) / n
CI_Y_diff <- Y_diff + z_score * sqrt(VY_diff)
```
:::

**Question to consider:** How does `CI_Y_diff` compare in width to
`CI_Y_1p`? Under what condition on the prediction surface would you expect
the difference estimator to help a lot, versus barely at all?

## Model-assisted (plug-in) estimate

Fit a simple linear model of `BA_ONF` on `BA_pred_NFI` using the sample,
verify the zero-mean-residual property, and use it to compute
$\hat{Y}_{plugin}$.

::: {.callout-tip collapse="true"}
## Hint 7 — Plug-in estimator

``` r
mod <- lm(BA_ONF ~ BA_pred_NFI, data = sample_plots)
mean(resid(mod))                 # should be ~0

R2 <- summary(mod)$r.squared
(1 - R2) * VY_1p                 # heuristic expected variance

Y_plugin  <- as.numeric(coefficients(mod)[1] +
  exhaustive_vector * coefficients(mod)[2])
VY_plugin <- var(mod$residuals) / n
CI_Y_plugin <- Y_plugin + z_score * sqrt(VY_plugin)
```
:::

**Question to consider:** `VY_plugin` here treats $\hat\beta_0$ and
$\hat\beta_1$ as if they were known constants rather than estimated
quantities. Does that make it exactly unbiased, asymptotically unbiased, or
biased, under uniform iid sampling?

## g-weight variance

Now account properly for the sampling variability of the regression
coefficients themselves, using a sandwich-type design-based covariance
estimator.

::: {.callout-tip collapse="true"}
## Hint 8 — g-weight variance

``` r
# Make a function
design_cov_matrix_beta <- function(mod, data){

  mf  <- model.frame(mod, data = sample_plots)
  y   <- model.response(mf)
  X   <- model.matrix(mod, data = sample_plots)
  res <- resid(mod)

  n_total <- nrow(X)
  As2inv    <- solve((t(X) %*% X) / n_total)
  theta_hat <- drop((y %*% X / n_total) %*% As2inv)

  middle_term <- (t(X) %*% ((res^2) * X)) / (n_total^2)
  cov_theta   <- As2inv %*% middle_term %*% As2inv
  
  return(cov_theta)
}

cov_B <- design_cov_matrix_beta(mod, data)
aux_vec <- c(1, as.numeric(exhaustive_vector))
Y_gw <- as.numeric(aux_vec %*% coef(mod))
g_var   <- drop(t(aux_vec) %*% cov_B %*% aux_vec)
CI_Y_gw <- Y_plugin + z_score * sqrt(g_var)
```
:::

**Question to consider:** How does `g_var` compare to `VY_plugin`? Why are they so similar?

::: {.callout-tip collapse="true"}
# Hint

The sample size is large, the point estimators are equivalent and both variance estimators
are design-consistent.

:::


## Alternative sythetic estimator with confidence intervals

We can borrow the design-based covariance estimate from the g-weight to 
fit a linear model **without** an intercept, in order to test a synthetic estimator
that has no guarantee to be unbiased via the zero mean residual property.

``` r

mod_no_intercept <- lm(BA_ONF ~ -1 + BA_pred_NFI, data = sample_plots)
mean(resid(mod_no_intercept))                 # should not be 0

cov_B_no_intercept <- design_cov_matrix_beta(mod_no_intercept)
aux_vec_no_intercept <- c(as.numeric(exhaustive_vector))
Y_syn2 <- as.numeric(aux_vec_no_intercept %*% coef(mod_no_intercept))
VY_syn2   <- drop(t(aux_vec_no_intercept) %*% cov_B_no_intercept %*% aux_vec_no_intercept)
CI_Y_syn2 <- Y_syn2 + z_score * sqrt(VY_syn2)

```

## Compare everything

Put the point estimates, variances, and confidence intervals for all four
estimators side by side.

```{r}
#| label: compare
#| eval: false

library(tidyverse)

estimates <- tibble(
  estimator = c(
    "One-phase", "Synthetic", "Synthetic2", "Difference",
    "Internal (Plug-in)", "Internal (g-weight)"
  ),
  estimate = c(Y_1p, Y_syn, Y_syn2, Y_diff, Y_gw, Y_plugin),
  variance = c(VY_1p, VY_syn, VY_syn2, VY_diff, VY_plugin, g_var),
  lower = c(
    CI_Y_1p[1], CI_Y_syn[1], CI_Y_syn2[1],
    CI_Y_diff[1], CI_Y_plugin[1], CI_Y_gw[1]
  ),
  upper = c(
    CI_Y_1p[2], CI_Y_syn[2], CI_Y_syn2[2],
    CI_Y_diff[2], CI_Y_plugin[2], CI_Y_gw[2]
  )
) %>%
  mutate(
    CI_width = upper - lower,
    type = if_else(
      estimator %in% c("Synthetic", "Synthetic2"),
      "Potentially biased",
      "Unbiased (at least asymptotically)"
    ),
    estimator = factor(estimator, levels = estimator)
  )

estimates


plot_estimates <- function(estimates, y_max = 24) {
  
  # Define Y-axis limits with margin
  y_min <- min(estimates$lower, na.rm = TRUE)
  margin <- 0.05 * (y_max - y_min)
  
  # Generate plot
  ggplot(estimates, aes(x = estimator, y = estimate, colour = type)) +
    geom_linerange(
      aes(ymin = lower, ymax = upper),
      linewidth = 1.2
    ) +
    geom_point(size = 3) +
    scale_colour_manual(
      values = c(
        "Potentially biased" = "red",
        "Unbiased (at least asymptotically)" = "blue"
      )
    ) +
    coord_cartesian(
      ylim = c(y_min - margin, y_max + margin)
    ) +
    labs(
      x = NULL,
      y = "Estimate",
      colour = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 30, hjust = 1),
      legend.position = "top"
    )
}

plot_estimates(estimates)

```

**Questions to consider:**

1. Which estimator has the smallest variance? Is that a reason to trust it
   the most?
2. Why does it not make sense to combine these intervals by simply taking their overlapping region?
3. Try refitting the synthetic2 model using only data plots in REGIONN == "AVANT-MONTS JURASSIENS" and generate the results again.
What happened to the bias now?  This is traditionally why "design-based" is considered the *gold standard* in inference for official statistics.

::: {.callout-tip collapse="true"}
## Hint: refit the model with this before re-running the previous estimate
```{r}
#| eval: false

mod_fit_locally <- lm(BA_ONF ~ BA_pred_NFI, data = sample_plots |> filter(REGIONN == "AVANT-MONTS JURASSIENS"))

# Note that the zero mean property only applies to the residuals in the training set.  
# It is lost here because we apply the model only in AVANT-MONTS JURASSIENS 
# and apply it to all other regions.  We can do this manually...

resids_manual <- sample_plots$BA_ONF - predict(mod_fit_locally, newdata = sample_plots)
mean(resids_manual)

cov_B_fit_locally <- design_cov_matrix_beta(mod_fit_locally)
aux_vec_fit_locally <- c(1, as.numeric(exhaustive_vector))
Y_syn2 <- as.numeric(aux_vec_fit_locally %*% coef(mod_no_intercept))
VY_syn2   <- drop(t(aux_vec_fit_locally) %*% cov_B_fit_locally %*% aux_vec_fit_locally)
CI_Y_syn2 <- Y_syn2 + z_score * sqrt(VY_syn2)
```
:::

# Model-assisted versus stratification

Recall that our dataset had different sampling intensities in some regions as indicated by the 
factor variable `surf_maille_fct`.  Let's see how effective this oversampling strategy was estimating
basal area in Doubs.

When you decide *before* selecting your sample that you will apply an **independent sampling design
in different domains** we can apply stratification.  This is in contrast with post-stratification
where we first select the sample according to a predetermined sampling design and apply an ANOVA
model — i.e. a linear model with categorical/indicator variables by strata  For uniform i.d.d. sampling
these methods can be shown to be asymptotoically equivalent.  We will compare them here.

The (pre-) stratification estimator is a weighted mean:

$$
\hat Y_{strat}
=
\sum_{h=1}^H p_h\,\bar Y_{s,h}
\qquad
\text{where}
\qquad
p_h=\frac{A_h}{A_T}
$$
where $A_h$ is the area of stratum $h$ and $\bar Y_{s,h}$ is the sample mean in $h$.

By definition that sampling between strata is independent, so we can easily sum the variances:
$$
\widehat{\mathbb{V}}[\hat Y_{strat}]=\sum_{h=1}^H p^2_h\cdot\widehat{\mathbb{V}}[\bar Y_{s,h}]=\sum_{h=1}^H p^2_h\cdot\left[\frac{1}{n_h(n_h-1)}\sum_{x\in s_h}(Y(x)-\bar Y_{s,h})^2\right]
$$
**Questions:** 

1. Why is $p_h$ get squared in the variance estimator?
2. In the lecture the point estimator for post-stratification was exactly the same
   but the variance estimator looked different.  Why can't we use the same variance estimator

::: {.callout-tip collapse="true"}
## Answer
The sampling designs are fundamentally different.  In stratification, the sample size in 
each stratum, $n_h$, is fixed, but in post-stratification it is random.  Furthermore, in the latter
the sample size in one stratum is correlated with the others -- more sample in one stratum means less in another
since $n$ is fixed.
:::

Try to implement the stratified estimator on the full dataset of field data in `doubs_ONF_plots`.  Hints are given
but try to do each step on your own before peeking:

**Step 1:**

Calculate $p_h$ using `doubs_intersection`.  It may be helpful to make a data.frame of `REGIONN` by `surf_maille_fct`
as follows:

```{r}
sampling_intensities <- doubs_ONF_plots |>
            select(REGIONN, surf_maille_fct) |>
            st_drop_geometry() |>
            distinct()
```

::: {.callout-tip collapse="true"}
## Hint: `merge(doubs_intersected, sampling_intensities)` and then `group_by(surf_maille_fct)` to sum `area_ha`.
```{r}
# Get total areas by sampling intensity variable "surf_maille_fct"
# and make into proportion
stratification_props <- doubs_intersected |> 
  merge(sampling_intensities) |>
  group_by(surf_maille_fct) |>
  mutate(strat_area = sum(area_ha)) |>
  st_drop_geometry() |>
  select(surf_maille_fct, strat_area) |>
  distinct() |>
  ungroup() |>
  mutate(strat_prop = strat_area/sum(strat_area))
  
stratification_props
```
:::

**Step 2:**

Merge $p_h$ to `doubs_ONF_plots`.

::: {.callout-tip collapse="true"}
# Hint
```{r}
# Merge to plot data
doubs_ONF_plots <- doubs_ONF_plots |> 
  merge(stratification_props, by = "surf_maille_fct", all.x = TRUE)
```
:::

**Step 3:**

Merge $p_h$ to `doubs_ONF_plots`.

::: {.callout-tip collapse="true"}
# Hint:   `st_drop_geometry()` and use `group_by(surf_maille_fct)` to calculate within strata means/variances.
```{r}
# Make (pre-) stratified estimate
strat_ests <- doubs_ONF_plots |> 
  st_drop_geometry() |>
  # filter(uniform_sample) |>
  group_by(surf_maille_fct) |>
  summarise(n_strat = n(),
            strat_prop = unique(strat_prop),
            Y_1p_h = mean(BA_ONF),
            VY_1p_h = (1/n_strat)*var(BA_ONF)
            )
```
:::

::: {.callout-tip collapse="true"}
# Hint: `ungroup()` and `summarise(...)` to sum the columns of strata means/variances
```{r}
# Make (pre-) stratified estimate
strat_ests <- doubs_ONF_plots |> 
  # filter(uniform_sample) |>
  st_drop_geometry() |>
  group_by(surf_maille_fct) |>
  summarise(n_strat = n(),
            strat_prop = unique(strat_prop),
            Y_1p_h = mean(BA_ONF),
            VY_1p_h = (1/n_strat)*var(BA_ONF)
            ) |>
  ungroup() |>
  summarise(Y_strat = sum(strat_prop*Y_1p_h),
            VY_strat = sum((strat_prop^2)*VY_1p_h)
            ) |>
  ungroup()
```
:::

**Step 4:**

Calculate the confidence intervals the same as before and add it to our `estimates` tibble.

::: {.callout-tip collapse="true"}
# Point estimate and variance with confidence interval
```{r}
CI_strat <- strat_ests$Y_strat + z_score*sqrt(strat_ests$VY_strat)
```
:::


::: {.callout-tip collapse="true"}
# Append to `estimates` tibble
```{r}
#| eval: false
estimates_appended <- estimates |>
  add_row(
    estimator = "Stratification (oversampled)",
    estimate = strat_ests$Y_strat,
    variance = strat_ests$VY_strat,
    lower = CI_strat[1],
    upper = CI_strat[2],
    CI_width = CI_strat[2]-CI_strat[1],
    type = "Unbiased (at least asymptotically)"
  )
  
estimates_appended
  
plot_estimates(estimates_appended)
```
:::

At the cost of 40% more sample plots, it kind of stinks as a stratification...  Why?


::: {.callout-tip collapse="true"}
# Hint:  Post-stratification is model-assisted estimation with an ANOVA...
```{r}
#| eval: false
# The associated model only explains 9% of the variance.
summary(lm(BA_ONF ~ surf_maille_fct, data = doubs_ONF_plots, weights = surf_maille))$r.squared
```
:::

# Model-assisted Small Area Estimation (SAE)

The main idea behind model-assisted small area estimation is to fit the model using the full sample to obtain the regression coefficient vector $\hat{\boldsymbol{\beta}}_s$ and its design-based variance-covariance matrix $\hat\Sigma_{\hat{\boldsymbol{\beta}}_s}$, and then use the fitted model to predict at sample locations within the small area $G$. The small area estimate can then be variance-estimated either using the plug-in estimator under the external model assumption or using the $g$-weight estimator. Two approaches are possible: 

1) de-bias the small area estimate using the mean residual in $G$, or 
2) include an indicator for $G$ in the model so that the mean residual in $G$ is zero by construction, and then use the $g$-weight variance estimator.

The general form of the first approach can be expressed as:

$$
\hat{Y}_{sae}=\bar{\mathbf Z}_G^\top\mathbf{\hat{\boldsymbol{\beta}}_s}+\frac{1}{n_G}\sum_{x\in {s_G}}\hat R(x),
\qquad\qquad
\hat{\mathbb V}[\hat{Y}_{sae}] = 
  \displaystyle
  \bar{\mathbf Z}_G^\top
  \hat\Sigma_{\hat{\boldsymbol{\beta}}_s}
  \bar{\mathbf Z}_G + \frac{1}{n_G}\widehat{\mathbb{V}}_{s_G}\left[\hat R(x)\right]
$$

The second approach is the same except that the second terms are removed because the mean residual term is zero in $G$ by construction.  Note that we could also do a plug-in approach under the external model assumption and the variance estimator would reduce to:

$$
\hat{\mathbb V}[\hat{Y}_{sae,external}]=\frac{1}{n_G}\widehat{\mathbb{V}}_{s_G}\left[\hat R(x)\right]
$$
As always, rules of thumb should be treated with caution, but experience and simulation studies suggest that these approaches can provide close-to-nominal coverage for areas with $n_G$ as small as approximately 5--6 observations. For smaller areas, a model-based approach is generally required and is beyond the scope of this tutorial.

Following the recommendation of Mandallaz (2013), confidence intervals can be improved by using the Student's $t$ distribution:
$$
\mathrm{CI}_{1-\alpha}(\hat{Y})
=
\hat{Y}
\pm
t_{n-p;\,1-\alpha/2}
\sqrt{\hat{V}(\hat{Y})}.
$$
Although the $t$ distribution formally relies on an approximately normal sampling distribution, this is rarely problematic in practice. Moreover, the $t$ distribution is asymptotically normal, so its use is consistent with the preceding CLT-based theory.

To illustrate the usage of the small area estimators use the following code to select some local publilc forests to target:

```{r}
#| eval: false

# leaflet needs WGS84 projection so change to that
doubs_wgs <- st_transform(doubs_public_forests, 4326)
placettes_wgs <- st_transform(doubs_ONF_plots, 4326)

# Interactive selection gadget
select_forests_app <- function(forests, points) {
  ui <- fluidPage(
    tags$head(tags$style(type="text/css", "#map {height: calc(100vh - 80px) !important;}")),
    leafletOutput("map"),
    actionButton("done", "Done & Return Selected Names", class = "btn-primary", style = "margin-top: 10px;")
  )
  
  server <- function(input, output, session) {
    selected_ids <- reactiveVal(character(0))
    
    output$map <- renderLeaflet({
      leaflet(forests) %>%
        addProviderTiles(providers$CartoDB.Positron) %>%
        addPolygons(
          layerId = ~llib_frt,
          fillColor = "lightgray",
          fillOpacity = 0.5,
          color = "gray",
          weight = 1,
          popup = ~llib_frt
        ) %>%
        addCircleMarkers(
          data = points,
          radius = 3,
          color = "blue",
          fillOpacity = 0.8
        )
    })
    
    observeEvent(input$map_shape_click, {
      click <- input$map_shape_click
      if (is.null(click$id)) return()
      
      current <- selected_ids()
      if (click$id %in% current) {
        current <- setdiff(current, click$id) # Unselect if clicked again
      } else {
        current <- c(current, click$id)       # Select on click
      }
      selected_ids(current)
      
      # Highlight selected features in red
      selected_sf <- forests[forests$llib_frt %in% current, ]
      leafletProxy("map") %>%
        clearGroup("highlight")
      
      if (nrow(selected_sf) > 0) {
        leafletProxy("map") %>%
          addPolygons(
            data = selected_sf,
            group = "highlight",
            fillColor = "red",
            fillOpacity = 0.7,
            color = "darkred",
            weight = 2
          )
      }
    })
    
    observeEvent(input$done, {
      stopApp(selected_ids())
    })
  }
  
  # Opens in a dedicated window with fixed dimensions
  runGadget(ui, server, viewer = dialogViewer("Forest Selector", width = 1000, height = 800))
}

# Launch the app
target_small_areas <- select_forests_app(doubs_wgs, placettes_wgs)

# Your areas you selected should be stored here
target_small_areas
```
You can extract the forests you selected using filter...
```{r}
#| eval: false
# You can extract them using filter...
sample_plots |> filter(llib_frt %in% target_small_areas)
```
...or just added an indicator variable with mutate
```{r}
#| eval: false
sample_plots |> mutate(in_target_area = llib_frt %in% target_small_areas)

```
Fit the model on the big sample to stabilize the estimate of the coefficients in the linear model and apply it with proper bias correction to your selected target area using a model-assisted SAE estimator.
