---
title: "Monte Carlo Interpretation"
author: "Alexander Massey"
format:
  revealjs:
    theme: default
    incremental: false
    slide-number: true
    width: 1200
    height: 900
    margin: 0.04
execute:
  echo: false
  warning: false
  message: false
---

```{=html}
<style>
.reveal h2{font-size:1.3em;}
.reveal .slides section img{max-height:78vh;width:auto;}
.notes-col{font-size:.6em;line-height:1.5;text-align:left;padding-left:.5em;}
.notes-col ul{list-style:none;padding-left:0;}
.swatch{display:inline-block;width:.8em;height:.8em;margin-right:.4em;vertical-align:middle;}
</style>
```

```{r}
#| include: false
knitr::opts_chunk$set(fig.width=7,fig.height=7,fig.align="center", dev="png",dpi=220,out.width="100%")
set.seed(42)
xlim<-c(-2,12)
ylim<-c(-2,12)
forest_xlim <- c(0,10)
forest_ylim <- c(0,10)
n_trees<-20
trees<-data.frame(x=runif(n_trees,-2,12),y=runif(n_trees,-2,12))
sample_point<-c(x=4.3,y=5.8)
r<-2.6
theta<-seq(0,2*pi,length.out=200)
selected<-sqrt((trees$x-sample_point["x"])^2+(trees$y-sample_point["y"])^2)<=r
draw_circle<-function(cx,cy,rad,...){
 lines(cx+rad*cos(theta),cy+rad*sin(theta),...)
}
```
## Sample unit is a point

:::: {.columns}
::: {.column width="75%"}
```{r}
par(mar=c(.5,.5,.5,.5))
plot(trees$x,trees$y,type="n",xlim=xlim,ylim=ylim,axes=FALSE,asp=1,xlab="",ylab="")
points(trees$x[!selected],trees$y[!selected],pch=17,col="gray60",cex=1.5)
points(trees$x[selected],trees$y[selected],pch=17,col="forestgreen",cex=1.8)
draw_circle(sample_point["x"],sample_point["y"],r,col="black",lwd=2)
points(sample_point["x"],sample_point["y"],pch=4,col="black",cex=2.2,lwd=3)
```
:::
::: {.column width="25%"}
<div class="notes-col">
A circle centered on the sample point selects trees inside the inclusion radius.
</div>
:::
::::

## Duality principle: draw inclusion rings around trees

:::: {.columns}
::: {.column width="75%"}
```{r}
par(mar=c(.5,.5,.5,.5))
plot(trees$x,trees$y,type="n",xlim=xlim,ylim=ylim,axes=FALSE,asp=1,xlab="",ylab="")
for(i in which(selected)){
 xx<-trees$x[i]+r*cos(theta)
 yy<-trees$y[i]+r*sin(theta)
 polygon(xx,yy,col=adjustcolor("lightgreen",alpha.f=.18),border=NA)
 lines(xx,yy,col="forestgreen",lwd=1.5,lty=2)
}
points(trees$x[!selected],trees$y[!selected],pch=17,col="gray60",cex=1.5)
points(trees$x[selected],trees$y[selected],pch=17,col="forestgreen",cex=1.8)
points(sample_point["x"],sample_point["y"],pch=4,col="black",cex=2.2,lwd=3)
```
:::
::: {.column width="25%"}
<div class="notes-col">
Overlap of the light-green inclusion rings illustrates density accumulation.
</div>
:::
::::

## Boundary adjustment on trees (not plots)

:::: {.columns}
::: {.column width="75%"}
```{r}
#| dev.args:
#|   bg: "transparent"
par(mar=c(.5,.5,.5,.5),xpd=NA)
plot(trees$x,trees$y,type="n",xlim=xlim,ylim=ylim,axes=FALSE,asp=1,xlab="",ylab="")
points(trees$x[!selected],trees$y[!selected],pch=17,col="gray60",cex=1.5)
points(trees$x[selected],trees$y[selected],pch=17,col="forestgreen",cex=1.8)
usr<-par("usr"); shade<-adjustcolor("grey93",alpha.f=.45)
rect(usr[1],usr[3],forest_xlim[1],usr[4],col=shade,border=NA)
rect(forest_xlim[2],usr[3],usr[2],usr[4],col=shade,border=NA)
rect(forest_xlim[1],usr[3],forest_xlim[2],forest_ylim[1],col=shade,border=NA)
rect(forest_xlim[1],forest_ylim[2],forest_xlim[2],usr[4],col=shade,border=NA)
rect(forest_xlim[1],forest_ylim[1],forest_xlim[2],forest_ylim[2],border="black",lwd=2)
dist<-pmin(trees$x-xlim[1],xlim[2]-trees$x,trees$y-ylim[1],ylim[2]-trees$y)
sel<-which(selected); border_i<-sel[which.min(dist[sel])]
xx<-trees$x[border_i]+r*cos(theta)
yy<-trees$y[border_i]+r*sin(theta)
clip(forest_xlim[1],forest_xlim[2],forest_ylim[1],forest_ylim[2])
polygon(xx,yy,col=adjustcolor("lightgreen",alpha.f=.45),border=NA)
do.call("clip",as.list(usr))
lines(xx,yy,col="forestgreen",lwd=2,lty=2)
points(trees$x[border_i],trees$y[border_i],pch=17,bg="white",col="forestgreen",cex=2,lwd=2)
```
:::
::: {.column width="25%"}
<div class="notes-col">
The gray mask indicates areas outside the forest while still allowing outside trees to remain visible.
</div>
:::
::::
