---
title: "Monte Carlo Interpretation"
author: "Alexander Massey"
format:
  beamer:
    theme: "Madrid"
    colortheme: "default"
    aspectratio: 169
    slide-level: 2
    keep-tex: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(fig.width=7,fig.height=7,fig.align="center", dev="png",dpi=220,out.width="100%")
set.seed(42)
xlim<-c(-2,12)
ylim<-c(-2,12)
forest_xlim <- c(0,10)
forest_ylim <- c(0,10)
n_trees<-20
trees<-data.frame(x=runif(n_trees,-2,12),y=runif(n_trees,-2,12))
sample_point<-c(x=4.3,y=5.8)
r<-2.6
theta<-seq(0,2*pi,length.out=200)
selected<-sqrt((trees$x-sample_point["x"])^2+(trees$y-sample_point["y"])^2)<=r
draw_circle<-function(cx,cy,rad,...){
 lines(cx+rad*cos(theta),cy+rad*sin(theta),...)
}
```

## Sample unit is a point

:::: {.columns}
::: {.column width="58%"}
```{r}
par(mar=c(0,0,0,0))
plot(trees$x,trees$y,type="n",xlim=xlim,ylim=ylim,axes=FALSE,asp=1,xlab="",ylab="")
points(trees$x[!selected],trees$y[!selected],pch=17,col="gray60",cex=1.5)
points(trees$x[selected],trees$y[selected],pch=17,col="forestgreen",cex=1.5)
draw_circle(sample_point["x"],sample_point["y"],r,col="black",lwd=2)
points(sample_point["x"],sample_point["y"],pch=4,col="black",cex=2.2,lwd=3)
```
:::
::: {.column width="42%"}
\footnotesize
\begin{itemize}
    \item A circular plot is centered on the \textbf{sample point} selects trees inside the inclusion radius.
    \item[] 
    \item Multiple inclusion radii are possible based on DBH thresholds.
    \item[]
    \item There are infinite points in the forest.
    \item[]
    \item \textbf{Finite population sampling theory} is commonly used but the problem is poorly posed.
    \item[]
    \item Circles do not tesselate! Angle count sampling makes no sense...
\end{itemize}
:::
::::

## \textbf{Workaround:} Draw same inclusion rings around trees

:::: {.columns}
::: {.column width="58%"}
```{r}
par(mar=c(0,0,0,0))
plot(trees$x,trees$y,type="n",xlim=xlim,ylim=ylim,axes=FALSE,asp=1,xlab="",ylab="")
for(i in which(selected)){
 xx<-trees$x[i]+r*cos(theta)
 yy<-trees$y[i]+r*sin(theta)
 polygon(xx,yy,col=adjustcolor("forestgreen",alpha.f=.25),border=NA)
 lines(xx,yy,col="forestgreen",lwd=1.5,lty=2)
}
points(trees$x[!selected],trees$y[!selected],pch=17,col="gray60",cex=1.5)
points(trees$x[selected],trees$y[selected],pch=17,col="forestgreen",cex=1.5)
points(sample_point["x"],sample_point["y"],pch=4,col="black",cex=2.2,lwd=3)
```
:::
::: {.column width="42%"}
\footnotesize
\begin{definition}[Duality principle]
trees in radius around point same as point in radius around trees.
\end{definition}
\begin{itemize}
    \item[]
    \item Overlap of the inclusion rings illustrates density accumulation at a point.
    \item[]
    \item Sample unit is the point; not the trees.
    \item[]
\end{itemize}

$x$
: point at plot center

:::
::::

## Boundary adjustment on trees (not plots)

:::: {.columns}
::: {.column width="58%"}
```{r}
#| dev.args:
#|   bg: "transparent"
par(mar=c(0,0,0,0),xpd=NA)
plot(trees$x,trees$y,type="n",xlim=xlim,ylim=ylim,axes=FALSE,asp=1,xlab="",ylab="")
dist<-pmin(trees$x-xlim[1],xlim[2]-trees$x,trees$y-ylim[1],ylim[2]-trees$y)
sel<-which(selected); border_i<-sel[which.min(dist[sel])]
points(trees$x[-border_i],trees$y[-border_i],pch=17,col="gray60",cex=1.5)
points(trees$x[border_i],trees$y[border_i],pch=17,col="forestgreen",cex=1.5)
usr<-par("usr"); shade<-adjustcolor("grey93",alpha.f=.45)
rect(usr[1],usr[3],forest_xlim[1],usr[4],col=shade,border=NA)
rect(forest_xlim[2],usr[3],usr[2],usr[4],col=shade,border=NA)
rect(forest_xlim[1],usr[3],forest_xlim[2],forest_ylim[1],col=shade,border=NA)
rect(forest_xlim[1],forest_ylim[2],forest_xlim[2],usr[4],col=shade,border=NA)
rect(forest_xlim[1],forest_ylim[1],forest_xlim[2],forest_ylim[2],border="black",lwd=2)
xx<-trees$x[border_i]+r*cos(theta)
yy<-trees$y[border_i]+r*sin(theta)
clip(forest_xlim[1],forest_xlim[2],forest_ylim[1],forest_ylim[2])
polygon(xx,yy,col=adjustcolor("forestgreen",alpha.f=.25),border=NA)
do.call("clip",as.list(usr))
lines(xx,yy,col="forestgreen",lwd=2,lty=2)
points(trees$x[border_i],trees$y[border_i],pch=17,bg="white",col="forestgreen",cex=1.5,lwd=2)
```
:::
::: {.column width="42%"}
\footnotesize
- Plots with centers outside forest (gray) are **not visited** for budget reasons.

- Requires boundary adjustment for **trees**.

\begin{definition}[Indicator variable]
\[
  I_i(x)=
  \begin{cases}
  1,&\text{if }x\in K_i,\\
  0,&\text{if }x\notin K_i.
  \end{cases}
\]
\end{definition}

\begin{definition}[Inclusion probability]
\[
  \pi_i = P[I_i(x)] = E[I_i(x)] =
  \frac{\lambda(K_i \cap F)}{\lambda(F)}
\]
\end{definition}

$K_i$
: inclusion ring around tree $i$

$\lambda(...)$
: surface area of ...

$F$
: forest domain of interest
:::
::::

## Local density surface from above

:::: {.columns}
::: {.column width="58%"}
```{r}
#| dev.args:
#|   bg: "transparent"
par(mar=c(0,0,0,0),xpd=NA)
plot(trees$x,trees$y,type="n",xlim=xlim,ylim=ylim,axes=FALSE,asp=1,xlab="",ylab="")
in_box <- (trees$x >= forest_xlim[1] & trees$x <= forest_xlim[2]) & (trees$y >= forest_ylim[1] & trees$y <= forest_ylim[2])
dist<-pmin(trees$x-xlim[1],xlim[2]-trees$x,trees$y-ylim[1],ylim[2]-trees$y)
sel<-which(selected); border_i<-sel[which.min(dist[sel])]
points(trees$x[!selected],trees$y[!selected],pch=17,col="gray60",cex=1.5)
points(trees$x[in_box],trees$y[in_box],pch=17,col="black",cex=1.5)
usr<-par("usr"); shade<-adjustcolor("grey93",alpha.f=.45)
rect(usr[1],usr[3],forest_xlim[1],usr[4],col=shade,border=NA)
rect(forest_xlim[2],usr[3],usr[2],usr[4],col=shade,border=NA)
rect(forest_xlim[1],usr[3],forest_xlim[2],forest_ylim[1],col=shade,border=NA)
rect(forest_xlim[1],forest_ylim[2],forest_xlim[2],usr[4],col=shade,border=NA)
rect(forest_xlim[1],forest_ylim[1],forest_xlim[2],forest_ylim[2],border="black",lwd=2)

for(i in which(in_box)){
xx<-trees$x[i]+r*cos(theta)
yy<-trees$y[i]+r*sin(theta)
clip(forest_xlim[1],forest_xlim[2],forest_ylim[1],forest_ylim[2])
polygon(xx,yy,col=adjustcolor("forestgreen",alpha.f=.25),border=NA)
do.call("clip",as.list(usr))
points(trees$x[in_box],trees$y[in_box],pch=17,bg="white",col="forestgreen",cex=1.5,lwd=2)  
}

for(i in which(in_box)){
xx<-trees$x[i]+r*cos(theta)
yy<-trees$y[i]+r*sin(theta)
clip(forest_xlim[1],forest_xlim[2],forest_ylim[1],forest_ylim[2])
# polygon(xx,yy,col=adjustcolor("forestgreen",alpha.f=.25),border=NA)
do.call("clip",as.list(usr))
lines(xx,yy,col="forestgreen",lwd=2,lty=2)
}
points(sample_point["x"],sample_point["y"],pch=4,col="black",cex=2.2,lwd=3)
```
:::
::: {.column width="42%"}
\footnotesize
\centering
\textbf{\underline{Surface is fixed; point is random}}
\begin{definition}[Local density surface]

\[
Y(x)=\frac{1}{\lambda(F)}\sum_{i=1}^N\frac{I_i(x)Y_i}{\pi_i}
\]

\end{definition}

$Y_i$
: measurement for tree $i$

$\pi_i$
: inclusion probability for tree $i$

$N$
: number of trees in forest $F$

$I_i(x)$
: indicator that tree $i$ is selected at $x$

$\lambda(F)$
: surface area of forest $F$


\begin{block}{Interpretation}

$Y(x)$ is the Horvitz--Thompson estimator with sample size one.

\end{block}
:::
::::
